---
title: "ERA Yield Stability"
authors: "Peter Steward & Todd Rosenstock"
email: "p.steward@cgiar.org; t.rosenstock@cgiar.org"
date: "`r Sys.Date()`"
output: html_vignette
vignette: >
  %\VignetteIndexEntry{ERA Yield Stability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup,include=F,eval=T,echo=T}
require(ERAg)
require(broom)
require(ggplot2)
require(metafor)
require(metan)
require(pbapply)
require(plyr)
require(sfsmisc)
require(MASS)
```

In this vignette we will explore how to plot and calculate absolute and relative outcome stability.

## Access the data
First we access the compiled ERA dataset included with the ERA package.

```{r Access ERA, echo=T}
knitr::kable(head(ERA.Compiled[,1:8], 5))
```

Our analyses of outcome stability are inspired and based on the publication *A global meta-analysis of yield stability in organic and conservation agriculture* by [Knapp *et al.* 2018](https://doi.org/10.1038/s41467-018-05956-1). As such we will be exploring the outcome subindicator crop yield in this vignette, which is a data rich vein in ERA.

## Rename higher level concepts

Many functions in the ERA package have been designed work with standardized column names for the higher level concepts of **practice**, **outcome** and **product** (or experimental units) found in the ERA organizational hierarchy.  

We can view the subordinates of these high level concepts by accessing the internal datasets `PracticeCodes`, `OutcomesCodes` and `EUCodes`.

```{r Show Practice Codes, echo=T}
knitr::kable(head(PracticeCodes[,1:6], 5))
```

Alternatively the organization of higher level concepts can also be viewed in the `ERAConcepts` list.
```{r Show ERA Concepts, echo=T}
ERAg::ERAConcepts
```

To apply standardized column names we need to select which organizational levels of the three higher level concepts to use using the `StandColName` function. The function help describes how to do this `?StandColNames`.

Crop yield is an outcome in the *subindicator* level of the outcome hierarchy, so we set the  `OLevel` parameter to `SI`. In general you will be using the *subindicator* level for stability analyses.

Regarding practices there are a large number of concepts at the *subpractice* level, so let's aggregate analysis up the *practice* level; we set the  `PLevel` parameter to `P`. 

The stability analysis does not explicitly consider products so it's not critical what organization level of prodct we choose; let's set the  `EULevel` parameter to `P`. 

```{r Run StandColName Function, echo=T}
StabData<-ERAg::StandColNames(Data=ERAg::ERA.Compiled,
              PLevel="P",
              OLevel="SI",
              EULevel="P"
              )
```

All we have done with the `StandColNames` function is to rename columns, for example the `Out.SubInd` column has been renamed `Outcome`.

To explore outcome x practice stability for specific products or product groups we could subset the dataset on the `products` columns, for example:

```{r Run Subset on Product, echo=T,eval=F}
StabData<-StabData[Product.Subtype=="Cereals"]
```

## Prepare the data

Next we need to prepare the data for use with the `StabCalc2` function. To do this we use the `PrepareStabData` function which labels multi-year observations (MYO) and calculates means and variances for outcomes within these MYOs. A multi-year observation is a pair of control and experimental treatments, in the same place and study with multiple observations of the same outcomes over time.

By default the `PrepareStabData` function subsets data to the crop yields outcome (code=101), but the `OutCodes` argument can be used to consider other outcomes. We can investigate outcome codes in the `OutcomeCodes` dataset.

```{r Show Outcome Codes, echo=T}
data.table(OutcomeCodes)[grep("Crop Yield",Subindicator),list(Subindicator,Code)]
```

```{r Prepare Stability Data, echo=T}
StabData<-ERAg::PrepareStabData(Data=StabData,OutCodes=101)
```

The stability analysis needs MYOs of at least 3 seasons to generate any statistics and the `StabCalc` function will filter any MYOs that do not meet this minimum data requirement. If you want to increase the thresholds for MYO length then you can filter on the `nryear` field of the `PrepareStabData` output. For example: 

```{r Subset Stability Data, echo=T,eval=F}
StabData<-StabData[nryears>=4]

```

## Analyse the data
Next we pass the `PrepareStabData` output to the `StabCalc2` function.

```{r Stability calc, echo=T, eval=T,include=F}
StabData<-ERAg::StabCalc2(Data=StabData,
          Do.Weight=T,
          Weight.by.Study=T,
          Rm.Out=T,
          Transform=T,
          Control=list(optimizer="optim",optmethod="Nelder-Mead",maxit=10000),
          Responses=c("lnRR","lnVR","lnCVR"),
          Use.acv = F)
```
## Plot the data
Now we use the `ERAStabPlot` function to plot the data, this function creates a list of plots, one for each combination of product x outcome.

```{r Stability Plots, echo =T}
StabPlots<-ERAgON::ERAStabPlot(Data=StabData,Robust=F,Intercept=F)
```
Each list element contains two elements, these are ggplots of 1) absolute yield variability (`lnVR`) vs. yield (`lnRR`) and 2) relative yield variability (`lnCVR`) vs. yield (`lnRR`).

```{r Example Plots, echo =T,fig.width=7,fig.height=6.5,warning=F}
plot(StabPlots$`Crop Yield.Agroforestry Pruning`$lnVR)
plot(StabPlots$`Crop Yield.Agroforestry Pruning`$lnCVR)
```
