---
title: "ERA-Large-Climate-Files"
author: "Peter Steward"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ERA-Large-Climate-Files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r packages,include=F,eval=F,echo=T}
require(ERAg)
require(piggyback)
```

This short vignette will demonstrate how to use the `piggyback` package to download some large climate datasets for ERA spatial locations.

## Basic methods
These large datasets contain daily or monthly climate information for each unique ERA location as identified by the `Site.Key` field in the `ERA.Compiled` dataset.  
An ERA location is a centroid point (represented by the `Latitude` and `Longitude` fields in the `ERA.Compiled` dataset) buffered to a radius of spatial uncertainty (using the value in `Buffer` field  of the `ERA.Compiled` dataset).  
Climate data for each buffered site are extracted for each temporal slice of climate data using functions in the `terra` package then averaged for the buffer.  

## ERA large climate datasets
### **`POWER.CHIRPS`** 
This dataset combines daily climate information from the [NASA POWER](https://power.larc.nasa.gov/) at 0.5&deg resolution and [CHIRPS 2.0](https://www.chc.ucsb.edu/data/chirps) dataset API at 0.05&deg  resolution. 

NASA POWER Solar parameters are derived from NASA's GEWEX/SRB release 3.0 archive (July 1, 1983 – Dec. 31, 2007) and NASA’s CERES FLASHFlux project (Jan. 1, 2008 – to within about 7-days of real time). Meteorological parameters are derived from the NASA's GMAO MERRA-2 assimilation model (Jan. 1, 1981 to within a few months of real time) plus GEOS-5.12.4 FP-IT (End of MERRA-2 to within several days of real time).  

The data cover the period 1983-07-01 to 2018-12-31.  

`POWER.CHIRPS` fields:  
1) **`Year`** = calendar year  
2) **`Day`** = julian day of year  
3) **`Latitude`** = latitude in decimal degrees  
4) **`Longitude`** = longitude in decimal degrees  
5) **`Solar.Rad`** = Insolation Incident on a Horizontal Surface - MJ/m^2/day (POWER)  
6) **`Pressure`** = Surface Pressure - kPa (POWER)  
7) **`Specific.Humid`** = Specific Humidity at 2 Meters - kg kg-1 (POWER)  
8) **`Humid`** = Relative Humidity at 2 Meters - % (POWER)  
9) **`Temp.Mean`** = Temperature at 2 Meters - C (POWER)  
10) **`Temp.Max`** = Maximum Temperature at 2 Meters - C (POWER)  
11) **`Temp.Min`** = Minimum Temperature at 2 Meters - C (POWER)  
12) **`WindSpeed`** = Wind Speed at 2 Meters - m/s (POWER)  
13) **`Buffer`** = radius of spatial uncertainty for location (m)  
14) **`Site.Key`** = unique ERA site key concatenating latitude, longitude and buffer  
15) **`Date`** = date as yyyy-mm-dd  
16) **`DayCount`** =  the number of days since "1900-01-01"  
17) **`Altitude`** = site altitude derived from the [ASTER Global Digital Elevation Model](10.5067/ASTER/ASTGTM.003) at 1" resolution
18) **`SRad.Sub`** = if this field does not equal `NA` then values were not available from the POWER dataset and have beeen subsituted from AgMERRA
19) **`ETo`** = FAO Penman-Monteith potential evapotranspiration calculated using the methods presented in http://www.fao.org/docrep/X0490E/x0490e06.htm - mm (POWER)  
20) **`Rain`** = rainfall - mm (CHIRPS)  

Note if you want to access the CHIRPS data only this is available in a smaller file called `CHIRPS.RData`.


### **`TERRA`**

This object contains climate and climatic water balance data from the [TERRACLIMATE](http://www.climatologylab.org/terraclimate.html) dataset. TERRACLIM data have monthly temporal resolution and a ~4-km (1/24th degree) spatial resolution. The data cover the period from 1958-2018.  

Data were extracted from climatology datasets found [here](http://thredds.northwestknowledge.net:8080/thredds/catalog/TERRACLIMATE_ALL/summaries/catalog.html).


`TERRA` fields:  
1) **`Mean`** = mean variable value  
2) **`Median`** = median variable value  
3) **`SD`** = standard deviation of variable value  
4) **`N`** = number of pixels within the ERA site buffer  
5) **`ID`** =  equivalent of `Site.Key` field, unique spatial identifier concatenating latitude, longitude and buffer  
6) **`Variable`** = TERRACLIM variable, see below for details of these  
7) **`Year`** = year  
8) **`Month`** = month  
9) **`LT.Mean`** = mean for the year and month across the period 1958-2018  
10) **`LT.sd`** = standard deviation for the year and month across the period 1958-2018  
11) **`Deviance`** = difference between the current observation and the long-term average mean (`Mean` - `LT.Mean`)  
  
`TERRA` variables:  
1) **`aet`** = Actual Evapotranspiration  
2) **`def`** = Climate Water Deficit  
3) **`pet`** = Reference evapotranspiration (ASCE Penman-Montieth)  
4) **`PDSI`** = Palmer Drought Severity Index  
5) **`soil`** = Soil Moisture  


### **`AgMERRA`**
This object contains climate information from the [AgMERRA Climate Forcing Dataset for Agricultural Modeling
](https://data.giss.nasa.gov/impacts/agmipcf/agmerra/). AgMERRA data have daily temporal resolution and 0.5 degree spatial resolution. The data cover the period from 1980-2010.  

AgMERRA fields:  
1) **`Site.Key`** = unique ERA site key concatenating latitude, longitude and buffer  
2) **`Latitude`** = latitude in decimal degrees  
3) **`Longitude`** = longitude in decimal degrees  
4) **`Buffer`** = radius of spatial uncertainty for location (m)  
5) **`Year`** = calendar year  
6) **`Day`** = julian day of year  
7) **`Temp.Mean`** = Temperature at 2 Meters - C   
8) **`Temp.Max`** = Maximum Temperature at 2 Meters - C  
9) **`Temp.Min`** = Minimum Temperature at 2 Meters - C  
10) **`Solar.Rad`** = Solar Radiation = MJ/m2/day (Short wave)  
11) **`WindSpeed`** = Wind Speed at 2 Meters - m/s  
12) **`RH.Max`** = relative humidity at time of max temp - %  
13) **`Rain`** = rainfall - mm
14) **`Date`** = date as yyyy-mm-dd  
15) **`DayCount`** =  the number of days since "1900-01-01"  

## How to access the datasets

First create a save directory, for example:  

```{r Create Dir, echo =TRUE,eval=F}
LargeDir<-"Large Files/"
if(!dir.exists(LargeDir)){
  dir.create(LargeDir,recursive = T)
}

```

Set the github repo name to `peetmate/ERAg` and the release version tag which holds the large files to `Beta_v0.0.1`:  

```{r Set parameters, echo =TRUE,eval=F}
REPO<-"peetmate/ERAg"
TAG<-"Beta_v0.0.1"
```

Use the `piggyback::pb_download` function to download the file from the repo.

You will need to set the `file` parameter to one of:  
1) `POWER.CHIRPS.RData`  
2) `TERRA.RData`  
3) `AgMERRA.RData`  

You should have been provided with a token to download the `ERAg` package, supply this same token to the `pb_download` functions `.token` parameter to access the repo (contact p.steward@cgiar.org if you need a token).

```{r download large file, echo =TRUE,eval=F}

Sys.setenv(GITHUB_TOKEN = "ghp_XXXXXXXXXXXXXXXXXXXXXXXX")

piggyback::pb_download(file= "AgMERRA.RData", 
            repo = REPO,
            tag = TAG,
            dest = LargeDir)
```

