---
title: "ERA-Merge-Biophysical"
author: "Peter Steward"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ERA-Merge-Biophysical}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Please familiarize yourself with the `ERA-Climate-Data` & `ERA-Large-Climate-Datasets` vignettes before reading this document.  

In this vignette we will be exploring how to connect ERA to soil and seasonal climate datasets.


### 1. Prepare analytical environment

```{r load packages, eval=F,echo=T}
  if(!require("pacman", character.only = TRUE)){
    install.packages("pacman",dependencies = T)
  }
  
  required.packages <- c("ERAg","data.table","miceadds","dplyr")
  p_load(char=required.packages,install = T,character.only = T)
    
```

```{r load packages real, eval=T,echo=F,include=F}
  require(ERAg)
  require(data.table)
  require(foreach)
```

#### Load climate data
In Section 2 of the `ERA-Climate-Data` vignette we saved the outputs of the `ClimateCalc` function to a folder `ERA Climate Analysis/` from which we will load the seasonal climate data we need to merge with ERA.

The output of `ClimateCalc` is saved in a folder named with a concatenation of the arguments provided to the function, in the 
 `ERA-Climate-Data` vignette they generate the folder name `422814143322303020152010130-3POWCHIFALSE0.6`.  
 
This climate analysis is also available from the `ERAg` git as a piggybacked file called `422814143322303020152010130-3POWCHIFALSE0.6.zip` see the `ERA-Large-Climate-Datasets.Rmd` for download instructions.


```{r  create dirs, eval=F,echo=T}
# Create a directory for the climate analysis
ClimDir<-"ERA Climate Analysis/422814143322303020152010130-3POWCHIFALSE0.6/"
if(!dir.exists(ClimDir)){
  dir.create(ClimDir)
}
```

Load saved seasonal climate data:
  
```{r  Load ClimateA, eval=F,echo=T}
ClimateA<-load(paste0(ClimDir,"ClimStatsA.RData"))
```

The `ClimateCalc` function also saved the processed input ERA dataset which contains the fields we need to link the climate data back to the ERA dataset. Load this dataset:

```{r  Load ERA, eval=F,echo=T}
ERA<-load(paste0(ClimDir,"Data.RData"))
```

#### Prepare iSDA Soils Data
[iSDA soils data](https://www.isda-africa.com/isdasoil/) are provided for each location in ERA.

```{r load soils,echo=T,eval=T}
Soils<-ERAgON::ERA_ISDA
Soils[,1:3]
```
For explanations of the soil variables in the `Variables` field see the
[Africa Soil and Agronomy Data Cube](https://gitlab.com/openlandmap/africa-soil-and-agronomy-data-cube) gitlab and/or the publication [Hengl, T., Miller, M.A.E., Križan, J. et al. (2021) African soil properties and nutrients mapped at 30 m spatial resolution using
two-scale ensemble machine learning. Sci Rep 11, 6130](https://doi.org/10.1038/s41598-021-85639-y).  

```{r soils show Variable names,echo=T,eval=T}
Soils[,unique(Variable)]
```

The soil variable names in `Variable` are bit a clunky (see above), let's simplify them by removing everything but the variable code from the name. We want to retain the depth information, so we create fields for upper and lower measurement depths.

```{r soils update Variable names,echo=T,eval=T}
Soils[grepl("0..200cm",Variable),Depth.Upper:=0
      ][grepl("0..200cm",Variable),Depth.Lower:=200
        ][,Variable:=gsub("_m_30m_0..200cm","",Variable)]

Soils[grepl("0..20cm",Variable),Depth.Upper:=0
      ][grepl("0..20cm",Variable),Depth.Lower:=20
        ][,Variable:=gsub("_m_30m_0..20cm","",Variable)]

Soils[grepl("20..50cm",Variable),Depth.Upper:=20
      ][grepl("20..50cm",Variable),Depth.Lower:=50
        ][,Variable:=gsub("_m_30m_20..50cm","",Variable)]

Soils[,unique(Variable)]

```

Two soil depths are provided for many variables, if you want to convert these to a single value the we can use a weighted mean based on soil depths. Alternatively you could provide your own weightings, perhaps you consider upper depths in the soil profile to be more important for crop yields than lower depths. You could also just consider the upper soil profile (in this case 0-20cm) if you feel greater depths are unimportant.

```{r soil weights, echo=T,eval=T}
Soils[,Weight:=Depth.Lower-Depth.Upper]

Soils<-unique(Soils[,list(Median=weighted.mean(Median,Weight),
                          Mean=weighted.mean(Mean,Weight),
                          SD=weighted.mean(SD,Weight),
                          Mode=Mode[Weight==max(Weight)]),
                    by=list(Variable,Site.Key)])
```

The `Soil` data.table is in long form, if we are to merge with an ERA dataset we will need to cast it to wide form where each site occupies a single row and each column is a soil variable x statistic.

```{r soils dcast, echo=T,eval=T}
colnames(Soils)

Soils<-dcast(Soils,Site.Key~Variable,value.var = c("Median","Mean","SD","Mode"))

colnames(Soils)
```

### 2. Merge climate & iSDA soil variables with ERA ####

The planting date used to estimate seasonal climate by the `CalcClimate` function is called `P.Date.Merge` in the processed ERA dataset and `PD.Used` in the `ClimateA[["Observed"]]` data. We use this field in combination with the location `Site.Key` and  crop `EU` fields to match the climate data to ERA.  

```{r merge MCode, echo=T,eval=F}
Clim.Data<-ClimateA[["Observed"]]

# Concatenate location, crop codes and planting date fields
Clim.Data[,MCode:=paste(ID,EU,PD.Used,sep="|")]
ERA[,MCode:=paste(Site.Key,EU,P.Date.Merge,sep="|")]
```

To estimate seasonal deviance for the variables in `ClimateA[["Observed"]]` we will need to match these data to the long-term averages in the `ClimateA[["LongTerm"]][["LT.Clim.Avg"]]` data.table. As there are no planting dates in the long-term average data we use growing season to match datasets.

```{r merge MCode LT, echo=T,eval=F}
LTAvg.Clim.Data<-ClimateA[["LongTerm"]][["LT.Clim.Avg"]]

# Concatenate location, crop codes and season fields
LTAvg.Clim.Data[,MCode.LT:=paste(ID,EU,Season)]
Clim.Data[,MCode.LT:=paste(ID,EU,Season)]

```

Now that we have made a key field, we use an lapply to append climate and soil data to ERA making a list entry for each value of `W.Name` in `Clim.Data`. `W.Name` indicates a calculation window for climate stats. Default values of `W.Name` are `Data` or `EcoCrop`, the former uses a season length based on reported or data estimated harvest dates the latter uses EcoCrop to estimate season length. Additional `W.Name` values present are custom season start and end dates supplied to `CalcClimate`.

Note that the `ClimateA[["LongTerm"]]` data are not directly comparable to the `Observed` data as they are usually derived from planting dates estimated using rainfall data. Whereas the `Observed` dataset typically uses reported planting and harvest dates to define a growing season. Despite methodological differences we consider the long-term average climate variables estimated from rainfall to be the best data available and can see no-other obvious pathway to obtaining better data on season deviance from "typical" conditions.

```{r merge lapply, echo=T,eval=F}
Merge Climate & Soil Data with ERA ####
ERA.Clim.Soil<-lapply(Clim.Data[,unique(W.Name)],FUN=function(X){
  
  # Subset seasonal climate data to calculation window X
  CData<-Clim.Data[W.Name==X]
  
  # Subset long-term climate averages to calculation window X
  LT.Clim.Data<-LTAvg.Clim.Data[W.Name==X]

  # Subset long-term climate averages for mean and median variables
  LT.Clim.Data.Mean<-LT.Clim.Data[Variable=="Mean"]
  LT.Clim.Data.Median<-LT.Clim.Data[Variable=="Median"]
  
  # Male an object containing the column names of climate variables for which we wish to calculate deviance
  Cols<-colnames(CData[,!c("Tmax.sd","Tmean.sd","PD.Used","W.Start","W.End","W.Name","ID","M.Year","Season","MCode","MCode.LT","EU","ETo.NA")])
  
  # Subtract long term average climate from the observed seasonal climate
  Mean.Diff<-CData[,..Cols]-LT.Clim.Data.Mean[match(CData[,MCode.LT],LT.Clim.Data.Mean[,MCode.LT])][,..Cols]
  Median.Diff<-CData[,..Cols]-LT.Clim.Data.Median[match(CData[,MCode.LT],LT.Clim.Data.Median[,MCode.LT])][,..Cols]
  
  # Rename columns to reflect they are now deviance from long-term mean or median values
  colnames(Mean.Diff)<-paste0(colnames(Mean.Diff),".Dev.Mean")
  colnames(Median.Diff)<-paste0(colnames(Median.Diff),".Dev.Median")
  
  # Join seasonal climate with deviance from long-term average tables
  CData<-cbind(CData,Mean.Diff,Median.Diff)
  
  # Match soils data to ERA using Site.Key field
  SoilData<-Soils[match(ERA[,Site.Key],Site.Key)]
  
  # Join ERA, Climate and Soils datasets removing any duplicated columns
  cbind(ERA, 
        CData[match(ERA$MCode,MCode),!c("MCode","EU","ID")], 
        SoilData[,!"Site.Key"])
  })

# Name list levels
names(ERA.Clim.Soil)<-unique(Clim.Data$W.Name)
```

```{r load vignette data, echo=F,eval=T,include=F}
ERA.Clim.Soil<-ERA.Clim.Soil.Vignette
```

We now have an ERA data.table for each level of `W.Name` with a large number of climate and soil fields appended. See the `?ClimateCalc` documentation for explantions of the different climate fields.

```{r show results, echo=T,eval=T}
names(ERA.Clim.Soil)

ERA.Clim.Soil[["EcoCrop"]][!is.na(ETo.sum)][c(1,30)]

```


